---
title: "comparative scWGCNA v1.0.0 workflow Gg_lumb_int with ctrl_int and poly_int"
author: "Fabio Sacher"
date: "01.03.2023"
output:
  html_document:
    toc: TRUE
    toc_float: TRUE
    df_print: paged
  html_notebook:
    fig_height: 7
    fig_width: 8
editor_options:
  chunk_output_type: inline
---

```{r}
library(scWGCNA)
library(Seurat)
library(gridExtra)
library(ggplot2)
library(magrittr)
library(dplyr)
source("~/spinal_cord_paper/scripts/scWGCNA_modplots.R")
```

```{r load-data}
# reference scWGCNA data
lumb_int_scWGCNA <- readRDS("~/spinal_cord_paper/output/Gg_lumb_int_scWGCNA_070223.rds")

# # load module annotation
# mod_annot <- read.delim("~/spinal_cord_paper/data/scWGCNA_Gg_lumb_int_module_annotation.txt",sep = ",")
# 
# knitr::kable(mod_annot)

# samples to compare on
Gg_samp <- c("Gg_ctrl_int", "Gg_poly_int")

# Gg = 1, Mm = 3
samples_sp <- c(1, 1)

# load samples
ctrl <- readRDS("~/spinal_cord_paper/data/Gg_ctrl_int_seurat_070223.rds")
poly <- readRDS("~/spinal_cord_paper/data/Gg_poly_int_seurat_070223.rds")


# concat and name list of samples
samples <- list(ctrl, poly)
names(samples) <- Gg_samp

#ortho table is gnames for it is only intraspec. comparisons
gnames<- modplots::gnames %>%
  dplyr::rename(gene = Gene.stable.ID) %>%
  dplyr::rename(name = Gene.name)

ortho <- gnames

```

## Pseudocells

```{r calc-pseudocells}

ps <- list()

for (i in names(samples)) {
  ps[[i]] <- calculate.pseudocells(samples[[i]], dims = 1:10)
}

```

## scWWGCNA compare

```{r scWGNA-compare, warning = FALSE}

# We then run the comparative analysis
Gg_lumb_int_comp <- scWGNA.compare(
          scWGCNA.data = lumb_int_scWGCNA,
          test.list = ps,
          test.names = Gg_samp,
          ortho = ortho, # not needed unless reference and tests have different gene names
          ortho.sp = samples_sp
          )

```

```{r module-frac}
knitr::kable(Gg_lumb_int_comp$misc$modulefrac)

# non-orthologous
Gg_lumb_int_comp$misc$geneslost[[1]] <- Gg_lumb_int_comp$misc$geneslost[[1]] %>%
  dplyr::left_join(gnames, by = "gene")
head(Gg_lumb_int_comp$misc$geneslost[[1]])
# non-expressed
Gg_lumb_int_comp$misc$geneslost[[2]] <- Gg_lumb_int_comp$misc$geneslost[[2]] %>%
  dplyr::left_join(gnames, by = "gene")
head(Gg_lumb_int_comp$misc$geneslost[[2]])

```

```{r gene-barplot, fig.height=10,fig.width=6}
# We can also plot these fractions as barplots, for each module we used for comparisons
scW.p.modulefrac.mod(Gg_lumb_int_comp)
``` 

Here, we can plot 4 different aspects, the zscore of the overall preservation, which according to the original authors of WGCNA, is understood as a threshold. Values under 2 mean no evidence of preservation in the test sample, over 10 mean strong evidence of preservation. Moreover, we can plot the median rank, which we can use to compare the preservation of the different modules. And, to go into details of the modules, we can plot the zscore of density and connectivity preservation.

```{r preservation, fig.height = 7, fig.width=15}
scW.p.preservation(scWGCNA.comp.data = Gg_lumb_int_comp,
                   to.plot=c("preservation", "median.rank"))

scW.p.preservation.mod(scWGCNA.comp.data = Gg_lumb_int_comp,
                       pt.style = c(24, 22),
                      to.plot=c("preservation", "median.rank"))
```

```{r density-connectivity, fig.height = 7, fig.width=15}
scW.p.preservation(scWGCNA.comp.data = Gg_lumb_int_comp,
                   to.plot=c("density", "connectivity"))

scW.p.preservation.mod(scWGCNA.comp.data = Gg_lumb_int_comp,
                       pt.style = c(24, 22),
                   to.plot=c("density", "connectivity"))
```


```{r pdf, warning=FALSE}

pdf('~/spinal_cord_paper/figures/Gg_lumb_int_comp_scWGCNA.pdf', width = 17)
scW.p.modulefrac.mod(Gg_lumb_int_comp)
scW.p.preservation(scWGCNA.comp.data = Gg_lumb_int_comp,
                   to.plot=c("preservation", "median.rank"))

scW.p.preservation.mod(scWGCNA.comp.data = Gg_lumb_int_comp,
                       pt.style = c(24, 22),
                   to.plot=c("preservation", "median.rank"))

scW.p.preservation(scWGCNA.comp.data = Gg_lumb_int_comp,
                   to.plot=c("density", "connectivity"))

scW.p.preservation.mod(scWGCNA.comp.data = Gg_lumb_int_comp,
                       pt.style = c(24, 22),
                   to.plot=c("density", "connectivity"))
dev.off()

```


```{r}
saveRDS(Gg_lumb_int_comp, "~/spinal_cord_paper/output/Gg_lumb_int_comp_scWGCNA.rds")
```

```{r}
# Date and time of Rendering
Sys.time()

sessionInfo()
```




