---
title: "Seurat workflow of the integrated day 10 data"
author: "Fabio Sacher"
date: "05.09.2025"
output:
  html_document:
    toc: TRUE
    toc_float: TRUE
    df_print: paged
  html_notebook:
    fig_height: 7
    fig_width: 8
editor_options:
  chunk_output_type: inline
---

```{r, setup, include=FALSE}
knitr::opts_chunk$set(
  comment = '', fig.width = 10, fig.height = 7
)
```

Render this script with > sbatch ~/spinal_cord_paper/scripts/Seurat_Gg_day_10_int_render.sh

# load the data

the following workflow is based on:
 - [https://github.com/satijalab/seurat/issues/1500] Cell cycle regression
 - [https://github.com/satijalab/seurat/issues/1836] General workflow
 - [https://github.com/satijalab/seurat/issues/2590] Integrate all features (although used in the integration script)

```{r libraries, warning = FALSE, message = FALSE, echo = T, results = 'hide'}
library(Seurat)
library(dplyr)
library(Rtsne)
library(RColorBrewer)
library(tidyr)
library(ggplot2)
library(gridExtra)
library(patchwork)
library(cowplot)
library(ggdendro)
library(ggrepel)
library(tibble)
```

```{r input, echo = T, results = 'hide', message=FALSE, warning=FALSE}

gnames <- modplots::gnames
# load the integrated seurat
my.int <- readRDS(file = "~/spinal_cord_paper/data/Gg_day10_integrated_050925.rds")

my.int@project.name <- "Gg_day10_int"

my.int
```

Date: `r format(Sys.Date(), "%d.%m.%y")`

# inspect

We order the plot by nCount_RNA to prevent overplotting of the last sample (default order is the group.by ident).

```{r inspection}
# have a look at the PC and the previously calculated UMAP
PCAPlot(
  my.int,
  group.by = "orig.ident",
  order = "nCount_RNA"
  )

DimPlot(
  my.int,
  reduction = "umap",
  group.by = "orig.ident",
  order = "nCount_RNA"
  )

```

# cell cycle scoring

Now we can score the cell cycle stage, using Seurats function. For this wee need the ortholog of the distributed cell stage marker lists. 

We do the CC.difference calculation at this stage - and not prior to integration - based on the suggestion from: [https://github.com/satijalab/seurat/issues/1500]

We add the scores of stages S and G2M, as well as the difference between them to the metadata with the names: S, G2M, CC.Difference

```{r , message=FALSE, warning=FALSE, echo = T, results = 'hide'}
# Load the orhtology table
ortho_gg_mm_v102 <- readRDS("~/spinal_cord_paper/data/ortho_gg_mm_v102.rds") 

colnames(ortho_gg_mm_v102) <- c(
  "GG_gene_ID",
  "GG_gene_Name",
  "MM_gene_ID",
  "MM_gene_Name",
  "ortho_conf",
  "homolog_type"
)

# A list of cell cycle markers, from Tirosh et al, 2015, is loaded with Seurat.  We can
# segregate this list into markers of G2/M phase and markers of S phase
s.genes <- ortho_gg_mm_v102 %>%
  dplyr::mutate(MM_gene_Name = toupper(MM_gene_Name)) %>%
  dplyr::filter(MM_gene_Name %in% cc.genes$s.genes) %>%
  dplyr::arrange(match(MM_gene_Name, cc.genes$s.genes)) %>%
  dplyr::pull(GG_gene_ID)

g2m.genes <- ortho_gg_mm_v102 %>%
  dplyr::mutate(MM_gene_Name = toupper(MM_gene_Name)) %>%
  dplyr::filter(MM_gene_Name %in% cc.genes$g2m.genes) %>%
  dplyr::arrange(match(MM_gene_Name, cc.genes$g2m.genes)) %>%
  dplyr::pull(GG_gene_ID) 

t0 <- Sys.time()
my.int <- CellCycleScoring(
  my.int,
  s.features = s.genes,
  g2m.features = g2m.genes,
  set.ident = TRUE
  )
paste0("Seurats CC scoring took ", Sys.time() - t0, " seconds to run.")

my.int$CC.Difference.seurat <- my.int$S.Score - my.int$G2M.Score
# view cell cycle scores and phase assignments
head(my.int[[]])

rm(t0, s.genes, g2m.genes)
```

## PCA plot of non-regressed data
```{r}
PCAPlot(my.int, group.by = "Phase")
```


# rescale integrated assay

We rescale the data to regress out the CC.difference

```{r scale.data-regress.cc.diff}
all_genes <- rownames(my.int)

my.int <- ScaleData(
  my.int,
  features = all_genes,
  vars.to.regress = "CC.Difference.seurat"
)
```

# PCA

```{r}
# Run the actual PCA (by default uses var.features of assay)
my.int <- RunPCA(my.int, verbose = F)

PCAPlot(my.int, group.by = "Phase")

# Which dimensions will we choose?
hist(my.int@reductions$pca@stdev^2, breaks = 500)

my.dimensions=1:20

```

# Dim Reduction

We will run the tSNE. Using the FFT-accelerated Interpolation-based t-SNE (FIt-SNE).
We run two, a normal tSNE, and an exaggerated tSNE, where clusters are tighter togheter. This is located under the xtsne name.

```{r tsne, message=FALSE, warning=FALSE, echo = T, results = 'hide'}

# Get the tSNE function
source("~/Software/FIt-SNE-master/fast_tsne.R")

my.tsne = fftRtsne(my.int@reductions$pca@cell.embeddings[,my.dimensions],
                   max_iter = 1500,
                   learning_rate =  round(dim(my.int)[2]/12),
                   initialization =(my.int@reductions$pca@cell.embeddings[,1:2]/my.int@reductions$pca@stdev[1])*0.0001,
                   perplexity_list = c(25, round(dim(my.int)[2]/100)),
                   fast_tsne_path="~/Software/FIt-SNE-master/bin/fast_tsne")

colnames(my.tsne) = c("tsne_1", "tsne_2")
rownames(my.tsne) = colnames(my.int)

my.int[["tsne"]] = CreateDimReducObject(embeddings = my.tsne, key = "tsne_", assay = DefaultAssay(my.int), global = T)

```

# Clustering

Here we perform the Louvain-jaccard clustering implemented in Seurat.
We can see the tree of clusters, to see how the clusters relate in the PCA space.   
We also use the tree, to check if any of the terminal pairs of sisters should be merged. This is determined based on a minimum of 20 DEs between the clusters.

```{r clusters, message=FALSE, warning=FALSE, echo = T, results = 'hide'}

# Find first the nearest neighbors
my.int <- FindNeighbors(object = my.int, dims=my.dimensions, verbose = F)

# Then the actual clusters
my.int <- FindClusters(object = my.int, resolution = 1, verbose = F, random.seed = 42)

# Check the tree of clusters, to see what's the relationship between them
my.int <- BuildClusterTree(my.int, dims = my.dimensions, verbose = F)
plot(Tool(object = my.int, slot = 'BuildClusterTree'))

# We are gonna check for DEs using the non integrated data. We are only gonna test genes that have variability, so we calculate variable genes
my.int <- FindVariableFeatures(my.int, assay = "RNA")


# Only the genes with variability > median
my.HVF <- HVFInfo(my.int, assay = "RNA")
my.HVF <- rownames(my.HVF)[which(my.HVF[,3] > (median(my.HVF[,3])))]
# We check for pairs of clusters, how mayn DEs they have. If less than n, we merge them
keep.check <- T
while (keep.check == T) {
  
  keep.check <- F
  # Check the tree of clusters, to see what's the relationship between them
  my.int <- BuildClusterTree(my.int, dims = my.dimensions, verbose = F)
  # Check only the terminal sisters
  to.check = ips::terminalSisters(my.int@tools$BuildClusterTree)
  for (i in to.check) {
    # DE between the sisters
    my.DE <- FindMarkers(my.int, i[1], i[2], test.use = "MAST", latent.vars = c("CC.Difference.seurat"),
                         min.pct = 0.25, verbose = T, assay = "RNA", features = my.HVF,) %>%
      dplyr::filter(abs(avg_log2FC) > 0.5) %>%
      dplyr::filter(p_val_adj < 0.05)
    
    # If less than 5, merge, and repeat
    if (dim(my.DE)[1] < 5) {
      cat(paste0(dim(my.DE)[1], " genes differentially expressed between clusters ",i[1]," and ",i[2]," merging \n"))
      my.int <- SetIdent(my.int, cells = WhichCells(my.int, idents = i[2]), value = i[1])
      keep.check <- T
    }
    print(i)
  }
}
rm(to.check, my.HVF, my.DE)

```

```{r tsne_again, fig.height = 12, fig.width = 12}

# renumber starting from 1
my.ID <- factor(
  Idents(my.int),
  levels= levels(Idents(my.int))[base::order(as.numeric(levels(Idents(my.int))))])
levels(my.ID) <- seq(length(levels(my.ID)))
Idents(my.int) <- my.ID
my.int[["seurat_clusters"]] <- my.ID

# Check again the clusters
dim1 <- DimPlot(my.int, reduction = "tsne", cols = rainbow(length(levels(my.ID))), label = T, label.size = 5, pt.size = 1)
dim1 <- DimPlot(my.int, reduction = "tsne", cols = rainbow(38), label = T, label.size = 5, pt.size = 1)
dim1
```

Plot the tree again

```{r renumbered tree}
# tree based on PCA dims
my.int <- BuildClusterTree(my.int, dims = my.dimensions, verbose = F)
plot(Tool(object = my.int, slot = 'BuildClusterTree'))
```

Now that we have the final reductions, we'll choose one and we look at the statistics of the cells. 

```{r ,fig.height=5, fig.width=10}

cand_genes <-  c("OLIG2", "SOX9", "TUBB3", "SHH", "SLC18A3", "GATA2", "PAX2",
                 "TLX3", "IGFBP7", "HBBA", "IFI30", "MSX2", "PLP1", "CDH5",
                 "ENO1", "GSTA4", "GBE1", "RSPO1", "LECT1")

plots <- list()

for (i in seq(cand_genes)) {
  plots[[i]] <- modplots::mFeaturePlot(my.int, my.features = c(cand_genes[i],cand_genes[i]), gnames = gnames)
}

```

# tSNE plots broad cluster

```{r}
annot <- read.csv("~/spinal_cord_paper/annotations/Gg_day10_int_cluster_annotation.csv")
  
clust_col <- read.csv("~/spinal_cord_paper/annotations/broad_cluster_marker_colors.csv")

# rename for left join
annot <- annot %>% 
  mutate(number = factor(number, levels = 1:nrow(annot))) %>% 
  rename(seurat_clusters = number) 
    
# add cluster annotation to meta data
my.int@meta.data <- my.int@meta.data %>% 
  tibble::rownames_to_column("rowname") %>% 
  left_join(annot, by = "seurat_clusters") %>% 
  tibble::column_to_rownames("rowname")
  
# colors to plot
broad_cols <- clust_col %>% 
  filter(broad_cluster %in% annot$broad) %>% 
  pull(color)
  
# plot
broad_tsne <- DimPlot(
    object = my.int,
    reduction = "tsne",
    group.by = "broad",
    cols = broad_cols,
    label = TRUE,
    label.size = 5
    ) +
    ggtitle(NULL)+
    theme_void() +
    NoLegend() +
    annotate(
      "text",
      label = "broad day10 int",
      x = 0,
      y = 0,
      angle = 45
    )

broad_tsne
```


```{r qplots, fig.height = 15, fig.width = 10, message=FALSE, warning=FALSE, echo = T}

# set and get dim.reduct embeddings
my.reduc <- "tsne"
emb <- data.frame(Embeddings(my.int, my.reduc))
colnames(emb) <- c("reduc_1", "reduc_2")

meta <- my.int[[]] %>%
  tibble::rownames_to_column("cell_ID")


my.plots = list()

my.plots[[1]] = ggplot(emb[meta[order(meta$nCount_RNA),]$cell_ID, ], aes(x = reduc_1, y = reduc_2)) +
    geom_point(aes(color=sort(meta$nCount_RNA)), size=1, alpha = 0.4, pch = 19) + 
    scale_colour_gradientn(colours = c("gray90","gray90","gray80","yellow", "orange", "red", "darkred", "darkred")) +
    theme_classic() + labs(colour="UMI count", x = paste0(my.reduc, "_1"), y = paste0(my.reduc, "_2"))

my.plots[[2]] = ggplot(emb[meta[order(meta$nFeature_RNA),]$cell_ID, ], aes(x = reduc_1, y = reduc_2)) +
    geom_point(aes(color=sort(meta$nFeature_RNA)), size=1, alpha = 0.4, pch = 19) + 
    scale_colour_gradientn(colours = c("gray90","gray90","gray80","yellow", "orange", "red", "darkred", "darkred")) +
    theme_classic() + labs(colour="gene count", x = paste0(my.reduc, "_1"), y = paste0(my.reduc, "_2"))

my.plots[[3]] = ggplot(emb[meta[order(meta$nCount_SCT),]$cell_ID, ], aes(x = reduc_1, y = reduc_2)) +
    geom_point(aes(color=sort(meta$nCount_RNA)), size=1, alpha = 0.4, pch = 19) + 
    scale_colour_gradientn(colours = c("gray90","gray90","gray80","yellow", "orange", "red", "darkred", "darkred")) +
    theme_classic() + labs(colour="UMI count (SCT)", x = paste0(my.reduc, "_1"), y = paste0(my.reduc, "_2"))

my.plots[[4]] = ggplot(emb[meta[order(meta$nFeature_SCT),]$cell_ID, ], aes(x = reduc_1, y = reduc_2)) +
    geom_point(aes(color=sort(meta$nFeature_RNA)), size=1, alpha = 0.4, pch = 19) + 
    scale_colour_gradientn(colours = c("gray90","gray90","gray80","yellow", "orange", "red", "darkred", "darkred")) +
    theme_classic() + labs(colour="gene count (SCT)", x = paste0(my.reduc, "_1"), y = paste0(my.reduc, "_2"))

my.plots[[5]] = ggplot(emb[meta[order(meta$percent.mt),]$cell_ID, ], aes(x = reduc_1, y = reduc_2)) +
    geom_point(aes(color=(sort(meta$percent.mt))), size=1, alpha = 0.4, pch = 19) + 
    scale_colour_gradientn(colours = c("gray90","gray80","yellow", "orange", "red", "darkred", "darkred")) +
    theme_classic() + labs(colour="log1p MT percent", x = paste0(my.reduc, "_1"), y = paste0(my.reduc, "_2"))

my.plots[[6]] = ggplot(emb[meta[order(meta$CC.Difference.seurat),]$cell_ID, ], aes(x = reduc_1, y = reduc_2)) +
    geom_point(aes(color=sort(meta$CC.Difference.seurat)), size=1, alpha = 0.4, pch = 19) + 
    scale_colour_gradientn(colours = c("gray90","gray90","gray80","yellow", "orange", "red", "darkred", "darkred")) +
    theme_classic() + labs(colour="Cell Cycle\nS-G2M", x = paste0(my.reduc, "_1"), y = paste0(my.reduc, "_2"))

my.plots[[7]] = ggplot(emb[meta[order(meta$percent.rb),]$cell_ID, ], aes(x = reduc_1, y = reduc_2)) +
    geom_point(aes(color=sort(meta$percent.rb)), size=2, alpha = 0.4, pch = 19) + 
    scale_colour_gradientn(colours = c("gray90","gray90","gray80","yellow", "orange", "red", "darkred", "darkred")) +
    theme_classic() + labs(colour="percent.rb")

grid.arrange(grobs=my.plots, ncol=2)

```


```{r sex-dim-plot}

female <- c("Gg_ctrl_2", "Gg_lumb_2", "Gg_poly_2")

my.int[[]] <- my.int[[]] %>% 
  mutate(sex = case_when(
    orig.ident %in% female ~ "female",
    TRUE ~ "male"
  ))

sex_tsne <- TSNEPlot(my.int, group.by = "sex", shuffle = TRUE, cols = c("pink","cyan"))
sex_tsne
```


```{r orig.ident-red.plot, fig.height = 10, fig.width = 15}

dim.orig.ident <- ggplot(emb, aes(x = reduc_1, y = reduc_2)) +
    geom_point(aes(color=meta$orig.ident[sample(x = seq(nrow(emb)), size = nrow(emb), replace = FALSE)]), size=0.8, alpha = 0.6, pch = 19) + 
    scale_colour_manual(values = rainbow(length(table(my.int@meta.data$orig.ident))) ) +
    theme_classic() + labs(colour="Dataset", x = paste0(my.reduc, "_1"), y = paste0(my.reduc, "_2"))

dim.orig.ident
```

# DE analysis

Here we do the differential expression analysis, and end up with the marker genes lists.
We can also see the marker gene dot plot, for the top 2 marker genes per cluster

```{r DE, fig.height=10, message=FALSE, warning=FALSE, echo = T}

# Find all the marker genes, with these thresholds MAST
semarkers = FindMarkers(my.int,
                        ident.1 = "female",
                        ident.2 = "male",
                        features = NULL, 
                        group.by = "sex",
                        only.pos = FALSE, 
                        min.pct = 0.25, 
                        logfc.threshold =  0.5,
                        latent.vars = c("CC.Difference.seurat"), 
                        test.use = "MAST", 
                        assay = "RNA", 
                        return.thresh = 0.05)

# We only keep the significant ones
marker_df <- semarkers %>%
  tibble::rownames_to_column("Gene.stable.ID") %>% 
  left_join(gnames, by = "Gene.stable.ID") %>% 
  mutate(cluster = case_when(
    avg_log2FC > 0 ~ "female",
    avg_log2FC < 0 ~ "male"
  )) 

# Take only the top 10

cols <- c(female = "pink",
          male = "cyan",
          ns = "grey")


volplot <- ggplot(data = marker_df,
                       aes(x = avg_log2FC,
                           y = -log10(p_val_adj),
                           label = Gene.name,
                      color = cluster
                       )) +
  geom_point() +
  ggrepel::geom_text_repel(color = "black")+
  scale_color_manual(values = cols) +
  ylab("-log10(padj)") +
  ylim(c(0,330)) +
  theme_bw()+
  ggtitle("Sig. marker genes")

volplot
```

```{r}
w_genes <- rtracklayer::readGFF("/scicore/home/tschoppp/GROUP/references/genomes/ENS_g6/10XGenome_ENSG6_extended_060420/genes/genes.gtf") %>% 
  filter(seqid == "W") %>% 
  filter(type == "gene") %>% 
  pull(gene_id)

no_w_marker_df <- marker_df[!marker_df$Gene.stable.ID %in% w_genes,]

no_W_volplot <-  ggplot(data = no_w_marker_df,
                   aes(x = avg_log2FC,
                       y = -log10(p_val_adj),
                       label = Gene.name,
                       color = cluster
                       )) +
  geom_point() +
  ggrepel::geom_text_repel(color = "black")+
  scale_color_manual(values = cols) +
  ylab("-log10(padj)") +
  theme_bw()+
  ylim(c(0,330)) +
  ggtitle("Sig. marker genes no W genes")

no_W_volplot

```


# save

```{r saving, message=FALSE, warning=FALSE, echo = T}

saveRDS(my.int, paste0("~/spinal_cord_paper/data/", my.int@project.name, "_seurat_",format(Sys.Date(), "%d%m%y"),".rds"))

write.table(marker_df, sep = "\t", row.names = T, col.names = T,
            file = paste0("~/spinal_cord_paper/data/", my.int@project.name, "_fullDE_female_vs_male_",format(Sys.Date(), "%d%m%y"),".txt"), quote = F)

pdf("~/spinal_cord_paper/figures/Gg_day10_int_broad_cluster_tsne.pdf", width = 10, height = 10)
broad_tsne
sex_tsne + ggtitle("Day10int sex (pink female/cyan male") + theme_void() + NoLegend()
dev.off()

pdf("~/spinal_cord_paper/figures/Gg_da10_int_sex_volplot.pdf")
volplot
no_W_volplot
dev.off()

```

```{r sessionInfo}
# Date and time of Rendering
Sys.time()

sessionInfo()
```
