---
title: "Devel_int WGCNA modules expression in NT D5, D7, and D10 samples"
author: "Fabio Sacher"
date: "04.08.2023"
output:
  html_document:
    toc: TRUE
    toc_float: TRUE
  html_notebook:
    fig_height: 7
    fig_width: 8
editor_options:
  chunk_output_type: inline
---

Render this report with ~/spinal_cord_paper/scripts/Gg_devel_scWGCNA_module_analysis_render.sh.

```{r setup}
library(Seurat)
library(WGCNA)
library(tidyr)
library(ggplot2)
library(stringr)
library(patchwork)
library(tidyverse)
library(cowplot)
library(pheatmap)
library(gridExtra)
source("~/Neuraltube/scripts/heatmap4.R")

```

# Load individual seurat and test WGCNA data

The individual data sets are the Day 5 (Gg_D05_ctrl_seurat_070323), Day 7 (Gg_D07_ctrl_seurat_070323), and Day 10 (Gg_ctrl_1_seurat_070323) chicken spinal cord sets. The test WGCNA data are the modules calculated on the integrated data set of all three stages.

```{r data-sets}
se_path <- c("Gg_D05_ctrl_seurat_070323",
             "Gg_D07_ctrl_seurat_070323",
             "Gg_ctrl_1_seurat_070323")

clust_col <- read.csv("~/spinal_cord_paper/annotations/broad_cluster_marker_colors.csv") %>% 
  rename(broad = broad_cluster) %>% 
  select(-marker)

```

Order of the broad clusters for plotting purposes.

```{r ordering}
broad_order <- c("progenitors",
      "FP",
      "RP",
      "FP/RP",
      "neurons",
      "OPC",
      "MFOL",
      "pericytes",
      "microglia",
      "blood",
      "vasculature"
      )
```


```{r seurat-objects-and-annotations}
my.ses <- list()
col_table <- list()
ord_levels <- list()

for (i in seq(se_path)) {
  # load the data sets
  my.se <- readRDS(paste0("~/spinal_cord_paper/data/", se_path[i], ".rds"))
  annot <- read.csv(list.files("~/spinal_cord_paper/annotations",
                               pattern = str_remove(se_path[i], "_seurat_\\d{6}"),
                               full.names = TRUE))
  
  if(length(table(annot$number)) != length(table(my.se$seurat_clusters))) {
     stop("Number of clusters must be identical!")
  }
  
  # rename for left join
  annot <- annot %>% 
    mutate(fine = paste(fine, number, sep = "_")) %>% 
    mutate(number = factor(number, levels = 1:nrow(annot))) %>% 
    rename(seurat_clusters = number) 
  
  # cluster order for vln plots
  ord_levels[[i]] <- annot$fine[order(match(annot$broad, broad_order))]
  
  # create index for color coding
  col_table[[i]] <- annot %>%
    left_join(clust_col, by = "broad") %>% 
    select(c("fine", "color"))
  
  # add cluster annotation to meta data
  my.se@meta.data <- my.se@meta.data %>% 
    rownames_to_column("rowname") %>% 
    left_join(annot, by = "seurat_clusters") %>% 
    mutate(fine = factor(fine, levels = annot$fine)) %>% 
    column_to_rownames("rowname")
  
  my.ses[[i]] <- my.se

}

names(my.ses) <- c("D05", "D07", "D10")
names(col_table) <- c("D05", "D07", "D10")
names(ord_levels) <- c("D05", "D07", "D10")

rm(my.se, annot)
```

```{r input data, echo=TRUE}
# The reference WGCNA data. We can have several, if we want to test many at the same time
WGCNA_data = list()
WGCNA_data[[1]] = readRDS("~/spinal_cord_paper/output/Gg_devel_int_scWGCNA_250723.rds")
my.wsub =list()
my.wsub[[1]]= c(1:22)

# the name of each sample, as they appear in my.files and in the metadata of the combined object
my.samplenames = c("D05", "D07", "D10")

# This is just to add a little bit more sense to the modules, so that we don't get just a color. Corresponds to WGCNA_data
my.modulenames = list()
my.modulenames[[1]] = c(1:22)

```

# Module gene correlation

Here, we do a correlation matrix / heatmap, to see which cell clusters group togheter. This helps us to make more detailed dotplots.  
This part of the script can still be used to compare several WGCNA datasets in parallel.  

```{r}
# broad cluster color table
all_col <- do.call(rbind, col_table) %>% 
  rownames_to_column("sample") %>% 
  mutate(sample = substr(sample, 1, 3)) %>% 
  mutate(sample_celltype = paste(sample, fine, sep = "_")) %>% 
  select(c("color", "sample_celltype", "sample"))
```

# Meta data

```{r meta-data}
#Get a dataframe with annotations for all the samples and colors we need.
my.metam <- list()

for (i in seq(my.ses)) {
  my.metam[[i]] <- my.ses[[i]][[]]
  rownames(my.metam[[i]]) <- paste0(rownames(my.metam[[i]]), "_", i)
}
my.metam <- do.call(rbind, my.metam)

my.metam$orig.ident <- str_replace_all(my.metam$orig.ident, pattern =  "Gg_D05_ctrl", "D05")
my.metam$orig.ident <- str_replace_all(my.metam$orig.ident, pattern =  "Gg_D07_ctrl", "D07")
my.metam$orig.ident <- str_replace_all(my.metam$orig.ident, pattern =  "Gg_ctrl_1", "D10")

# my.metam$sample_celltype = paste0(substr(my.metam$orig.ident,7,9),"_",my.metam$seurat_clusters)
my.metam$sample_celltype = paste0(my.metam$orig.ident, "_", my.metam$fine)

my.metam <- my.metam %>% 
  tibble::rownames_to_column(var = "cell_ID") %>%
  dplyr::left_join(all_col, by = "sample_celltype") %>%
  tibble::column_to_rownames(var = "cell_ID")


# get sample colors
my.colsm = c("grey", "grey30", "black")
names(my.colsm) <- c("D05", "D07", "D10")
 
```


# Average expression data

```{r}

cell_table <- my.metam %>% 
  rownames_to_column("cell_ID") %>% 
  select("sample_celltype", "cell_ID")

avg.mod.eigengenes <- WGCNA_data[[1]]$sc.MEList$averageExpr %>% 
  rownames_to_column("cell_ID") %>% 
  left_join(cell_table, by = "cell_ID") %>% 
  column_to_rownames("cell_ID")
  

# add metadata

avg.mod.eigengenes.mean <- avg.mod.eigengenes %>%
  group_by(sample_celltype) %>%
  summarise_all("mean") %>%
  column_to_rownames("sample_celltype")

```

## spearman correlation heatmap

### annotations

```{r annot-list}

#The colors for the samples and clusters. First the closest to the heatmap. Add a white space to easy the eye and make less confussing
my.heatcols <- as.matrix(data.frame(
    sample= as.character(my.colsm[match(all_col$sample, names(my.colsm))]),
    "." = "white",
    cluster= as.character(all_col$color[match(rownames(avg.mod.eigengenes.mean), all_col$sample_celltype)]))
    )


# names and colors for the heatmap annotation
annot_name <- data.frame(
  "Celltypes" = all_col$sample_celltype,
  "Sample"    = all_col$sample,
  row.names = all_col$sample_celltype)
  

annot_module <- data.frame(
  "Module" = colnames(avg.mod.eigengenes)[1:22],
  row.names = colnames(avg.mod.eigengenes)[1:22]
)

pheat_col_table <- do.call(rbind, col_table) %>% 
  rownames_to_column("sample") %>% 
  mutate(sample = substr(sample, 1,3)) %>% 
  mutate(fine = paste(sample, fine, sep = "_"))

# match color table with annotation
pheat_col_table <- pheat_col_table[match(annot_name$Celltypes, pheat_col_table$fine),]

annot_col <- list(
  Celltypes = pheat_col_table$color,
  Sample = c(D05 = "#A4A4A4",
             D07 = "#515151",
             D10 = "#000000"),
  Module = str_remove(colnames(avg.mod.eigengenes)[1:22], "AE")
  )

names(annot_col[[1]]) <- annot_name$Celltypes
names(annot_col[[3]]) <- colnames(avg.mod.eigengenes)[1:22]

```

```{r,  fig.height=10, fig.width=13}
heat_col <- colorRampPalette(colors = c("dodgerblue4","dodgerblue", "white", "red", "darkred"))

#Calculate the distance 1-cor, and then calculate dendograms to cluster the clusters
my.hclust = hclust(as.dist(1-cor(t(avg.mod.eigengenes.mean), method = "spearman")))

# lower limit to scale color bar
low_limit <- 100 - abs(round(range(cor(t(avg.mod.eigengenes.mean), method = "spearman"))[1]*100))

pheatmap(cor(t(avg.mod.eigengenes.mean), method = "spearman"),revC = T,
                 main = "spearman correlation of average module eigengene expression by cluster\n Dendrogram based on 'distance 1-cor'",
                 fontsize = 8,  
                 color = heat_col(200)[low_limit:200],
                 show_colnames = F,
                 cluster_rows = my.hclust,
                 cluster_cols = my.hclust,
                 treeheight_row = 0, 
                 annotation_col = annot_name,
                 annotation_colors = annot_col,
                 annotation_legend = F,
                 border_color = NA
                 )

colbar <- c(min(cor(t(avg.mod.eigengenes.mean), method = "spearman")),
            max(cor(t(avg.mod.eigengenes.mean), method = "spearman")))

#Calculate the distance 1-cor, and then calculate dendograms to cluster the clusters
my.hclust = as.dendrogram(hclust(as.dist(1-cor(t(avg.mod.eigengenes.mean), method = "spearman"))))

#Plot!
pdf("~/spinal_cord_paper/figures/Fig_1_module_gene_heatmap_spearman.pdf", width = 11, height = 11)
heatmap.4(cor(t(avg.mod.eigengenes.mean), method = "spearman"),
          col = colorRampPalette(c("dodgerblue4","dodgerblue", "white", "red", "darkred"))(n = 1000),
          trace="none",
          lhei=c(0.2,1.6),
          lwid=c(0.2,1.6),
          scale = "none",margins = c(10,10),
          cexCol = 0.75,
          revC = TRUE,
          ColSideColors = my.heatcols,
          RowSideColors = t(my.heatcols[,3:1]),Rowv = my.hclust, Colv = my.hclust)


pdf("~/spinal_cord_paper/figures/Fig_1_module_gene_heatmap_color_key_spearman.pdf", width = 10, height = 10)
# heatmap since color key is missing in first heatmap
heatmap.4(rbind(colbar, colbar),
          col = colorRampPalette(c("dodgerblue4","dodgerblue", "white", "red", "darkred"))(n = 1000),
          trace="none",
          lhei=c(0.2,1.6),
          lwid=c(0.2,1.6),
          scale = "none",margins = c(10,10),
          cexCol = 0.75)

```

## heatmap of module pseudobulk average expression

```{r,  fig.height=15, fig.width=10}
# module colors
my.colcols = as.matrix(names(table(WGCNA_data[[1]]$dynamicCols)))

htmp <- pheatmap(as.matrix(avg.mod.eigengenes.mean),
         fontsize = 8,
         scale = "column",
         color = colorRampPalette(c("#4d2d87","#7c55a3", "white", "#ed9921", "#895d25"))(n = 1000),
         annotation_row = annot_name,
         annotation_col = annot_module,
         annotation_colors = annot_col,
         annotation_legend = F,
         border_color = NA)

module_order <- htmp[["tree_col"]]$order

pdf("~/spinal_cord_paper/figures/Fig_1_devel_module_v_clusters_heatmap.pdf", width = 7, height = 10)
pheatmap(as.matrix(avg.mod.eigengenes.mean),
         fontsize = 8,
         scale = "column",
         color = colorRampPalette(c("#4d2d87","#7c55a3", "white", "#ed9921", "#895d25"))(n = 1000),
         annotation_row = annot_name,
         annotation_col = annot_module,
         annotation_colors = annot_col,
         annotation_legend = F,
         border_color = NA)


```



# Load the integrated data set

The integrated data set on which the WGCNA is calculated. We plot it, split by the original cell types from the three original samples. 

```{r dimplots, fig.width=10}
# This is a file of all the combined mouse datasets, normalized and such.
my.sec = readRDS("~/spinal_cord_paper/data/Gg_devel_int_seurat_250723.rds")

identical(rownames(my.metam), colnames(my.sec))

my.metam$sample_celltype <- factor(my.metam$sample_celltype, levels = all_col$sample_celltype)
#Set the identities of the integrated data, to the annotated clusters
my.sec = SetIdent(my.sec, value = my.metam$sample_celltype)

p1 <- DimPlot(
  my.sec,
  reduction = "tsne",
  label = TRUE,
  repel = TRUE,
  cols = all_col$color,
  split.by = "orig.ident"
  ) + 
  NoLegend()

p1

pdf("~/spinal_cord_paper/figures/Devel_split_tsne.pdf", height = 7, width = 15)
#Plot split tsne
p1
```

# Avg. module exp. by stage

tSNE DimPlots showing the average expression of each module by stage.

```{r AE, message = FALSE, fig.width=12, fig.height=40}

for (i in seq(my.ses)) {
  # prepare average expression table
  tmp <- avg.mod.eigengenes[,1:22] %>%
    tibble::rownames_to_column("cell_ID") %>%
    dplyr::filter(grepl(paste0("_", i, "$"), cell_ID)) %>%
    dplyr::mutate(cell_ID = stringr::str_remove_all(cell_ID, paste0("_", i))) %>%
    tibble::column_to_rownames("cell_ID")
  
  identical(rownames(tmp), colnames(my.ses[[i]]))
  # add meta data to the seurat objects
  my.ses[[i]] <- AddMetaData(my.ses[[i]], tmp)
}

#max and min expression per module (column max)
mod_max <- apply(avg.mod.eigengenes[,1:22], MARGIN = 2, FUN = max)[module_order]
mod_min <- apply(avg.mod.eigengenes[,1:22], MARGIN = 2, FUN = min)[module_order]

modplots <- list()
modplots[[1]] <- list()
modplots[[2]] <- list()
modplots[[3]] <- list()

modules_in_order <- colnames(tmp)[module_order]

# plot the modules split to the stages
for (i in seq(my.ses)) {
  for (j in seq(ncol(tmp))) {
  
    modplots[[i]][[j]]  <- FeaturePlot(
      my.ses[[i]], order = TRUE,
      features = modules_in_order[j],
      reduction = "tsne"
      ) +
      ggtitle(stringr::str_remove(modules_in_order[j],"^AE")) +
      scale_color_gradient(low="ivory2", high=substring(modules_in_order[j], 3), #colors in the scale
                 limits=c(mod_min[j], mod_max[j])) #same limits for plots

    
    }
}

full_plot <- c(modplots[[1]], modplots[[2]], modplots[[3]])

gridExtra::grid.arrange(grobs = full_plot, ncol = 3, as.table = FALSE)

pdf("~/spinal_cord_paper/figures/Fig_2_devel_modules_AE_plots.pdf", width = 12, height = 70)
gridExtra::grid.arrange(grobs = full_plot, ncol = 3, as.table = FALSE)

pdf("~/spinal_cord_paper/figures/Supp_Fig_2_modules_darkgreen_AE_plots.pdf", width = 13, height = 4)
(full_plot[[9]] + full_plot[[31]] + full_plot[[53]]) + plot_layout(ncol = 3, guides = "collect")
```

# AE over time

We plot the average expression of each module in the three stages and the 5 broad cell type clusters present in all 3 stages.

```{r AE-over-time, fig.height=7, fig.width=12}
# module annotations
mod_annot <- read.csv("~/spinal_cord_paper/annotations/Gg_devel_int_scWGCNA_module_annotation.csv") %>%
  dplyr::mutate(module = str_replace_all(module, "\\d{1,2}\\_", "AE"))

meta <- list()

for (i in seq(my.ses)) {
  meta[[i]] <- my.ses[[i]]@meta.data %>%
    tibble::rownames_to_column("cell_ID") %>%
    dplyr::mutate(cell_ID = paste0(cell_ID, "_", i)) %>%
    dplyr::select(c("cell_ID", "broad"))
}

meta <- do.call(rbind, meta)

# mean average expression by stage
mean_AE <- avg.mod.eigengenes[,1:22] %>%
  tibble::rownames_to_column("cell_ID") %>%
  dplyr::mutate(stage = stringr::str_sub(cell_ID, -1)) %>%
  dplyr::mutate(stage = factor(stage, levels = c(1:3), labels = c("D05", "D07", "D10"))) %>%
  dplyr::left_join(meta, by = "cell_ID") %>%
  tibble::column_to_rownames("cell_ID") %>%
  tidyr::unite("stage_cl", stage, broad, sep = "_") %>%
  dplyr::group_by(stage_cl) %>%
  dplyr::summarise_each(mean) %>%
  dplyr::ungroup() %>%
  gather(key="module", value = "AE", -stage_cl) %>%
  dplyr::left_join(mod_annot[, c(1,3)], by = "module") %>%
  tidyr::separate("stage_cl", c("stage", "broad"), sep = "_", remove = FALSE) 

labels_dotplot <- stringr::str_remove(modules_in_order, "^AE")
names(labels_dotplot) <- modules_in_order

plot_clusters <- c("progenitors", "neurons", "RP", "FP", "pericytes", "OPC", "MFOL", "microglia", "blood")

mean_mod <- ggplot(data = mean_AE,
  aes(
    x = stage,
    y = AE,
    color = factor(broad, levels = plot_clusters),
    group = broad,
    label = annotation
    )
  ) +
  geom_line() +
  geom_point() +
  scale_color_manual(values = clust_col$color[match(plot_clusters, clust_col$broad)]) +
  theme_bw() +
  # facet wrap with reordered factors
  facet_wrap(vars(factor(module, levels = unique(mean_AE$module)[module_order])),
             scales = "free_y",
             nrow = 4,
             ncol = 6) +
  labs(color = "broad") +
  ggtitle("Average module expression by stage")

plotly::ggplotly(mean_mod)

pdf("~/spinal_cord_paper/figures/Fig_2_mean_mod_AE.pdf", height = 7, width = 12)
#Plot split tsne
mean_mod
```

# VlnPlots of avg. module exp. by stage and seurat cluster 

## colored by module

```{r average-module-expression-per-cluster, fig.width=10, fig.height=20}

# reorder seurat clusters
for (i in seq(my.ses)) {
  my.ses[[i]]$seurat_clusters <- factor(
  my.ses[[i]]$seurat_clusters,
  levels = levels(my.ses[[i]]$seurat_clusters)[as.integer(str_extract(ord_levels[[i]], "\\d{1,2}$"))]
  )

}

vplots <- list()

for (i in seq(my.ses)) {
  
  mods <- colnames(my.ses[[i]][[]])[grep("^AE",colnames(my.ses[[i]][[]]))]
  
  vplots[[i]] <- VlnPlot(
        my.ses[[i]],
        features = mods[module_order],
        group.by = "seurat_clusters",
        stack = TRUE, flip = TRUE,
        cols = substring(mods, 3)[module_order]) +
    theme(legend.position = "none") +
    geom_hline(yintercept = 0, lty = "dashed")
  
}

vplots[[1]]
vplots[[2]]
vplots[[3]]

pdf("~/spinal_cord_paper/figures/Fig_2_AE_by_cluster_modcol.pdf", height = 20, width = 10)
vplots[[1]]
vplots[[2]]
vplots[[3]]
```

## colored by cell type

```{r vln-by-cluster, fig.width=10, fig.height=20}
clust_col <- read.csv("~/spinal_cord_paper/annotations/broad_cluster_marker_colors.csv")

vplots_id <- list()

for (i in seq(my.ses)) {
    
  mods <- colnames(my.ses[[i]][[]])[grep("^AE",colnames(my.ses[[i]][[]]))]
  
  vplots_id[[i]] <- VlnPlot(
    my.ses[[i]],
    features = mods[module_order],
    group.by = "seurat_clusters",
    stack = TRUE, flip = TRUE,
    fill.by = "ident",
    cols = col_table[[i]]$color[as.integer(str_extract(ord_levels[[i]], "\\d{1,2}$"))]) +
    theme(legend.position = "none") +
    geom_hline(yintercept = 0, lty = "dashed")
}

vplots_id[[1]]
vplots_id[[2]]
vplots_id[[3]]

pdf("~/spinal_cord_paper/figures/Fig_2_AE_by_cluster_clucol.pdf", height = 20, width = 10)
vplots_id[[1]]
vplots_id[[2]]
vplots_id[[3]]
```

# MN modules

For the figures we specifically select the two MN modules and plot them as Vln plots.

```{r, fig.width=20, fig.height=5}
tmp <- avg.mod.eigengenes 
  
identical(rownames(tmp), colnames(my.sec))
# add meta data to the int seurat object
my.sec <- AddMetaData(my.sec, tmp)
my.sec <- AddMetaData(my.sec, my.metam[c("fine", "sample_celltype")])

custom_order <- c(paste(names(ord_levels)[1], ord_levels[[1]], sep = '_'),
                  paste(names(ord_levels)[2], ord_levels[[2]], sep = '_'),
                  paste(names(ord_levels)[3], ord_levels[[3]], sep = '_'))

my.sec$sample_celltype <- factor(
  my.sec$sample_celltype,
  levels = custom_order
  )

vln_ind <- VlnPlot(my.sec,
                   features = c("AEdarkred", "AElightgreen", "AEdarkgreen"),
                   group.by = "sample_celltype",
                   stack = TRUE,
                   flip = TRUE,
                   cols = c("darkred", "lightgreen", "darkgreen"),pt.size = 1
                  ) +
    NoLegend()

vln_ind
pdf("~/spinal_cord_paper/figures/Fig_2_AE_selected_mod.pdf", height = 22, width = 20)
vln_ind
```

```{r}
VlnPlot(
    my.sec,
    features = mods[module_order],
    group.by = "sample_celltype",
    stack = TRUE, flip = TRUE,
    cols = substring(mods, 3)[module_order]) +
    theme(legend.position = "none") +
    geom_hline(yintercept = 0, lty = "dashed")



pdf("~/spinal_cord_paper/figures/Fig_2_AE_by_cluster_integrated_data.pdf", height = 20, width = 30)
VlnPlot(
    my.sec,
    features = mods[module_order],
    group.by = "sample_celltype",
    stack = TRUE, flip = TRUE,
    cols = substring(mods, 3)[module_order]) +
    theme(legend.position = "none") +
    geom_hline(yintercept = 0, lty = "dashed")


```

```{r}
# Date and time of Rendering
Sys.time()

sessionInfo()
```

