---
title: "Devel_int WGCNA modules expression in NT D5, D7, and D10 samples"
author: "Fabio Sacher"
date: "04.08.2023"
output:
  html_document:
    toc: TRUE
    toc_float: TRUE
  html_notebook:
    fig_height: 7
    fig_width: 8
editor_options:
  chunk_output_type: inline
---

```{r library}
library(tidyverse)
library(Seurat)
library(ggrepel)
library(gridExtra)
library(pheatmap)
library(cowplot)
library(soilphysics)
```


# TF subset

```{r devel-WGCNA-tf-subset}
# TF table
tfs <- read.delim("~/../GROUP/references/genomes/TFs/Gallus_gallus_TF_AnimalTFDB3_edited.txt")
# gene names
gnames <- modplots::gnames

# Devel int scWGCNA
myWGCNA <- readRDS("~/spinal_cord_paper/output/Gg_devel_int_scWGCNA_250723.rds")

dev_modules <- lapply(myWGCNA[["module.genes"]], function(x) x[x %in% tfs$Ensembl])
names(dev_modules) <- paste0(names(myWGCNA[["modules"]]),"_")

# TF subset of devel modules 
mod_tf <- data.frame(Gene.stable.ID = unlist(dev_modules)) %>% 
  rownames_to_column("module") %>% 
  left_join(gnames, by = "Gene.stable.ID") %>% 
  mutate(module = str_remove(module, "_\\d{0,3}$")) %>% 
  mutate(module = str_remove(module, "^\\d{1,3}_")) %>% 
  mutate(tf_member = paste0(module, "_",Gene.stable.ID))
```

# Average module expression

```{r avg_module-exp}
# module average expression
avg_mod_exp <- myWGCNA[["sc.MEList"]][["averageExpr"]] 
colnames(avg_mod_exp) <- str_remove(colnames(avg_mod_exp), "^AE") 
```

# Module order and manual thresholds

The thresholds are manually set!

```{r}
# module order
module_order <- c("lightgreen","darkred","turquoise","brown","cyan",
                  "purple","midnightblue","lightcyan","blue","salmon",
                  "lightyellow","pink","royalblue","darkgreen","tan",
                  "yellow","grey60","greenyellow","green","red",
                  "magenta","black")


# elbowbplots of avg module expression
names <- c("black","blue","brown","cyan","darkgreen","darkred",
           "green","greenyellow","grey60","lightcyan","lightgreen",
           "lightyellow","magenta","midnightblue","pink","purple",
           "red","royalblue","salmon","tan","turquoise","yellow")


```

```{r elbowplots_model, fig.height= 4, fig.width=8}
fit_elbow <- list()
model_exp <- list()
fitted_values <- list()
coef_list <- list()

# shift values by min scaled expression (so all values are > 0)
shift_exp <- data.frame(apply(avg_mod_exp, 2, function(x) x - min(x)))

modules_to_plot <- rev(c("lightgreen","darkred","purple","lightcyan",
                     "blue","pink","royalblue","darkgreen","tan","yellow"))


b = c(0.003, 0.02, 0.0007, 0.0007,0.0003,0.005,0.01,0.01,0.001,0.003)
c = c(1.3,0.7,2.3,1.7,2.2,1.4,1.6,1.6,1.6,1.6)

for (i in seq(modules_to_plot)) {
  module <- modules_to_plot[i]
  
  toplot <-  shift_exp %>% 
    arrange(desc(.data[[module]])) %>% 
    mutate(rank = as.numeric(row_number())) %>% 
    select(!!sym(module),rank) %>% 
    dplyr::rename("mod_exp" = !!sym(module))
  
  # fit exponential curve
  model_exp[[i]] <- nls(mod_exp ~ a * (1 + b * rank) ^ -c, data = toplot, start = list(a = max(toplot$mod_exp), b = b[i], c = c[i]))
  
  coef_list[[i]] <- coef(model_exp[[i]])
  
  fitted_values[[i]] <- predict( model_exp[[i]], newdata = toplot)

  fit_elbow[[i]] <- ggplot(toplot, aes(x = rank, y = (mod_exp))) +
    geom_line(col = module) +
    geom_line(aes(y = fitted_values[[i]])) +
    scale_x_continuous(breaks = seq(1, 15000, by = 14999))+
    ggtitle(module) +
    theme(
      axis.text.y = element_blank(),
      axis.ticks.length.y = unit(0, "cm"),
      panel.grid.major = element_blank(),
      panel.grid.minor = element_blank())

}

names(fit_elbow) <-(modules_to_plot)

elbow_fit <- grid.arrange(grobs = fit_elbow[modules_to_plot], ncol = 5, nrow = 2)
```

```{r}

pd_x <- list()
spline <- list()

for (i in seq(modules_to_plot)) {
  a <- coef_list[[i]][1]
  b <- coef_list[[i]][2]
  c <- coef_list[[i]][3]
  f <- function(x) a*(1+b*x)^-c
  # get a x value by perpendicular distance from secant line
  pd_x[[i]] <- maxcurv(x.range = c(1,5000), f, method = "pd", col = modules_to_plot[i])
  # use x limit from above in spline to obtain local maximum curvature
  spline[[i]] <- maxcurv(x.range = c(1,5000), f, method = "spline", x0ini = pd_x[[i]]["x0"], col = modules_to_plot[i])
}

names(pd_x) <- modules_to_plot
names(spline) <- modules_to_plot


```

```{r}
elbowplots_thres <- list()
# only the modules with clear elbows, on off states
mod_to_plot <- avg_mod_exp[modules_to_plot]

exp_min <- apply(mod_to_plot, 2, min)

adj_thresholds <- rep(0,length(exp_min))

for (i in seq(ncol(mod_to_plot))) {
  module <- modules_to_plot[i]
  
  toplot <-  avg_mod_exp %>% 
    arrange(desc(.data[[module]])) %>% 
    mutate(rank = as.numeric(row_number())) %>% 
    select(!!sym(module),rank) %>% 
    dplyr::rename("mod_exp" = !!sym(module))
  
  # since the data was shifted by min(avg_mod_exp) for model fitting
  # we have to add the min (negative) to the threshold
  adj_thresholds[i] <- unname(spline[[module]]$y0)+unname(exp_min[module])
  
  elbowplots_thres[[i]] <- ggplot(toplot, aes(x = rank, y = mod_exp)) +
    geom_hline(yintercept =  adj_thresholds[i], lty = "dashed") +
    geom_line(col = module) +
    scale_x_continuous(breaks = seq(1, 15000, by = 14999))+
    ggtitle(module) +
    theme(
      axis.text.y = element_blank(),
      axis.ticks.length.y = unit(0, "cm"),
      panel.grid.major = element_blank(),
      panel.grid.minor = element_blank())

}

names(elbowplots_thres) <- modules_to_plot
names(adj_thresholds) <- modules_to_plot

elbowplots_thres_overall <- grid.arrange(grobs = elbowplots_thres[modules_to_plot], ncol = 5, nrow = 2)
```

# elbow plots of module expression

Lines indicate detection threshold.

```{r elbowplots, fig.height= 6, fig.width=8}
elbowplots <- list()

thresholds <- data.frame(module = colnames(avg_mod_exp))

tresh_df <- data.frame(adj_thresholds) %>% 
  rownames_to_column("module") %>% 
  right_join(thresholds) %>% 
  column_to_rownames("module")
 
for (i in seq(ncol(avg_mod_exp))) {
  module <- colnames(avg_mod_exp)[i]
  
  toplot <-  avg_mod_exp %>% 
    arrange(desc(.data[[module]])) %>% 
    mutate(rank = as.numeric(row_number())) %>% 
    select(!!sym(module),rank) %>% 
    dplyr::rename("mod_exp" = !!sym(module))
  
  elbowplots[[i]] <- ggplot(toplot, aes(x = rank, y = mod_exp)) +
    geom_hline(yintercept = tresh_df[module, 1], lty = "dashed") +
    geom_line(col = module) +
    scale_x_continuous(breaks = seq(1, 15000, by = 14999))+
    ggtitle(module) +
    theme(
      axis.text.y = element_blank(),
      axis.ticks.length.y = unit(0, "cm"),
      panel.grid.major = element_blank(),
      panel.grid.minor = element_blank())

}

names(elbowplots) <- colnames(avg_mod_exp)

elbow_overall <- grid.arrange(grobs = elbowplots[rev(module_order)], ncol = 6, nrow = 4)
```


# TF vs module expression correlations

```{r tf-expression}
# seurat object to extrace TF expression from
my.se <- readRDS("~/spinal_cord_paper/data/Gg_devel_int_seurat_250723.rds")

exp_data <- my.se[["RNA"]]$data

tf_exp <- exp_data[rownames(exp_data) %in% tfs$Ensembl, ]
tf_exp <- tf_exp[!rowSums(tf_exp) == 0,]
# transform
tf_exp <- t(tf_exp)

```

## pearson

```{r pearson_cor, fig.width= 8, fig.height=6}
# Calculate Pearson correlation between each TF and module (apply takes 10 seconds)
ps_correlations <- apply(tf_exp, 
                         MARGIN = 2, 
                         function(x) apply(avg_mod_exp,
                                           MARGIN = 2,
                                           function(y) cor(x, y, method = "pear")))

# Convert correlations to a data frame for easier analysis

ps_cor_long <- ps_correlations %>% 
  data.frame() %>% 
  rownames_to_column("module") %>% 
  mutate(module = factor(module, levels = module_order)) %>% 
  gather(key = "Gene.stable.ID", value = "corr", (2:(ncol(ps_correlations)+1))) %>% 
  left_join(gnames) %>% 
  mutate(tf_member = paste0(module, "_",Gene.stable.ID)) %>%
  mutate(member = case_when(
    tf_member %in% mod_tf$tf_member ~ "yes", 
    TRUE ~ "no"))

range(ps_cor_long$corr)

ps_correlations %>% 
  t() %>% 
  data.frame() %>% 
  summarise_all(min)

ps_correlations %>% 
  t() %>% 
  data.frame() %>% 
  summarise_all(max)

ps_cor_boxplot <- ps_cor_long %>% 
  mutate(outlier = case_when(
    abs(corr) > 0.5 ~ Gene.name,
    TRUE ~ NA
  )) %>% 
  ggplot(aes(x = module, y = corr, color = module, label = outlier)) +
  geom_boxplot() +
  scale_color_manual(values = module_order) +
  ylim(c(-1,1)) +
  geom_hline(yintercept = c(-0.5,0.5), lty = "dashed") +
  theme(axis.text.x = element_text(angle = 90, hjust = 1),
        legend.position = "none") +
  ggrepel::geom_text_repel(color = "black", max.overlaps = 20) +
  ggtitle("pearson corr TF vs module expression")


ps_cor_boxplot

ps_cor_filt <- ps_cor_long %>% 
  filter(abs(corr) > 0.5)
```

## spearman

```{r spearman, fig.width= 8, fig.height=6}
# Calculate Pearson correlation between each TF and module (apply takes 100 seconds)
sm_correlations <- apply(tf_exp, 
                         MARGIN = 2, 
                         function(x) apply(avg_mod_exp,
                                           MARGIN = 2,
                                           function(y) cor(x, y, method = "spearman")))

# Convert correlations to a data frame for easier analysis

sm_cor_long <- sm_correlations %>% 
  data.frame() %>% 
  rownames_to_column("module") %>% 
  mutate(module = factor(module, levels = module_order)) %>% 
  gather(key = "Gene.stable.ID", value = "corr", (2:(ncol(sm_correlations)+1))) %>% 
  left_join(gnames) %>% 
  mutate(tf_member = paste0(module, "_",Gene.stable.ID)) %>%
  mutate(member = case_when(
    tf_member %in% mod_tf$tf_member ~ "yes", 
    TRUE ~ "no"))

range(sm_cor_long$corr)

sm_correlations %>% 
  t() %>% 
  data.frame() %>% 
  summarise_all(min)

sm_correlations %>% 
  t() %>% 
  data.frame() %>% 
  summarise_all(max)

sm_cor_boxplot <- sm_cor_long %>% 
  mutate(outlier = case_when(
    abs(corr) > 0.5 ~ Gene.name,
    TRUE ~ NA
  )) %>% 
  ggplot(aes(x = module, y = corr, color = module, label = outlier)) +
  geom_boxplot() +
  scale_color_manual(values = module_order) +
  ylim(c(-1,1)) +
  geom_hline(yintercept = c(-0.5,0.5), lty = "dashed") +
  theme(axis.text.x = element_text(angle = 90, hjust = 1),
        legend.position = "none") +
  ggrepel::geom_text_repel(color = "black", max.overlaps = 20) +
  ggtitle("spearman corr TF vs module expression")

sm_cor_boxplot

sm_cor_filt <- sm_cor_long %>% 
  filter(abs(corr) > 0.5)
```

```{r tf_intesection}
intersect(sm_cor_filt$Gene.name, ps_cor_filt$Gene.name)
```

# Cell to cluster lookup table

```{r}
se_path <- c("Gg_D05_ctrl_seurat_070323",
             "Gg_D07_ctrl_seurat_070323",
             "Gg_ctrl_1_seurat_070323")

# clust_col <- read.csv("~/spinal_cord_paper/annotations/broad_cluster_marker_colors.csv") %>% 
#   rename(broad = broad_cluster) %>% 
#   select(-marker)

broad_order <- c("progenitors",
                 "FP",
                 "RP",
                 "FP/RP",
                 "neurons",
                 "OPC",
                 "MFOL",
                 "pericytes",
                 "microglia",
                 "blood",
                 "vasculature"
)

clust_col <- read.csv("~/spinal_cord_paper/annotations/broad_cluster_marker_colors.csv") %>% 
  rename(broad = broad_cluster) %>% 
  select(-marker)

my.ses <- list()
col_table <- list()
ord_levels <- list()

for (i in seq(se_path)) {
  # load the data sets
  my.se <- readRDS(paste0("~/spinal_cord_paper/data/", se_path[i], ".rds"))
  annot <- read.csv(list.files("~/spinal_cord_paper/annotations",
                               pattern = str_remove(se_path[i], "_seurat_\\d{6}"),
                               full.names = TRUE))
  
  if(length(table(annot$number)) != length(table(my.se$seurat_clusters))) {
    stop("Number of clusters must be identical!")
  }
  
  # rename for left join
  annot <- annot %>% 
    mutate(fine = paste(fine, number, sep = "_")) %>% 
    mutate(number = factor(number, levels = 1:nrow(annot))) %>% 
    rename(seurat_clusters = number) 
  
  # create index for color coding
  col_table[[i]] <- annot %>%
    left_join(clust_col, by = "broad") %>% 
    select(c("fine", "color"))
  
  # cluster order for vln plots
  ord_levels[[i]] <- annot$fine[order(match(annot$broad, broad_order))]
  
  # add cluster annotation to meta data
  my.se@meta.data <- my.se@meta.data %>% 
    rownames_to_column("rowname") %>% 
    left_join(annot, by = "seurat_clusters") %>% 
    mutate(fine = factor(fine, levels = annot$fine)) %>% 
    column_to_rownames("rowname")
  
  my.ses[[i]] <- my.se
  
}

names(my.ses) <- c("D05", "D07", "D10")
names(col_table) <- c("D05", "D07", "D10")
names(ord_levels) <- c("D05", "D07", "D10")

rm(my.se, annot)

#Get a dataframe with annotations for all the samples and colors we need.
my.metam <- list()

for (i in seq(my.ses)) {
  my.metam[[i]] <- my.ses[[i]][[]]
  rownames(my.metam[[i]]) <- paste0(rownames(my.metam[[i]]), "_", i)
}
my.metam <- do.call(rbind, my.metam)

my.metam$orig.ident <- str_replace_all(my.metam$orig.ident, pattern =  "Gg_D05_ctrl", "D05")
my.metam$orig.ident <- str_replace_all(my.metam$orig.ident, pattern =  "Gg_D07_ctrl", "D07")
my.metam$orig.ident <- str_replace_all(my.metam$orig.ident, pattern =  "Gg_ctrl_1", "D10")

# my.metam$sample_celltype = paste0(substr(my.metam$orig.ident,7,9),"_",my.metam$seurat_clusters)
my.metam$sample_celltype = paste0(my.metam$orig.ident, "_", my.metam$fine)


cell_table <- my.metam %>% 
  rownames_to_column("cell_ID") %>% 
  select("sample_celltype", "cell_ID")

# broad cluster color table
all_col <- do.call(rbind, col_table) %>% 
  rownames_to_column("sample") %>% 
  mutate(sample = substr(sample, 1, 3)) %>% 
  mutate(sample_celltype = paste(sample, fine, sep = "_")) %>% 
  select(c("color", "sample_celltype", "sample"))

rm(my.metam, my.ses)
```

# module detection rate

calulate detection rate per individual clusters based on the thresholds set above.

```{r detection-rate}

trsh_df <- data.frame(module = names(adj_thresholds),
                      thresholds = adj_thresholds)

detect_df <- mod_to_plot %>% 
  rownames_to_column("cell_ID") %>% 
  gather(key = "module", value = "avg_mod_exp", 2:11) %>% 
  left_join(trsh_df, by = "module") %>% 
  mutate(detection = case_when(
    avg_mod_exp > thresholds ~ TRUE,
    TRUE ~ FALSE
  )) %>% 
  left_join(cell_table, by = "cell_ID")

detect_rate <- detect_df %>% 
  mutate(key = paste(sample_celltype, module, sep = "/")) %>% 
  group_by(key) %>% 
  summarise(detection_rate = mean(detection == TRUE)) 

detect_rate %>% 
  separate_wider_delim(key, delim = "/", names = c("sample_celltype", "module")) %>% 
  arrange(desc(detection_rate)) %>%  
  mutate(rank = row_number()) %>% 
  ggplot(aes(x = rank, y = detection_rate)) +
  geom_line() 
```

# mean module expression per module

```{r}
# names and colors for the heatmap annotation
annot_name <- data.frame(
  "Celltypes" = all_col$sample_celltype,
  "Sample"    = all_col$sample,
  row.names = all_col$sample_celltype)
  

annot_module <- data.frame(
  "Module" = colnames(mod_to_plot),
  row.names = colnames(mod_to_plot)
)

pheat_col_table <- do.call(rbind, col_table) %>% 
  rownames_to_column("sample") %>% 
  mutate(sample = substr(sample, 1,3)) %>% 
  mutate(fine = paste(sample, fine, sep = "_"))

# match color table with annotation
pheat_col_table <- pheat_col_table[match(annot_name$Celltypes, pheat_col_table$fine),]

annot_col <- list(
  Celltypes = pheat_col_table$color,
  Sample = c(D05 = "#A4A4A4",
             D07 = "#515151",
             D10 = "#000000"),
  Module = colnames(mod_to_plot)
  )

names(annot_col[[1]]) <- annot_name$Celltypes
names(annot_col[[3]]) <- colnames(mod_to_plot)
```


```{r mean-clust-exp, fig.height= 10, fig.width=8}
cluster_mod_exp <- mod_to_plot %>% 
  rownames_to_column("cell_ID") %>% 
  left_join(cell_table, by = "cell_ID") %>% 
  column_to_rownames("cell_ID") %>%
  group_by(sample_celltype) %>%
  summarise_all("mean") %>% 
  column_to_rownames("sample_celltype")

# heatmap to get the row and col orders
htmp <- pheatmap(as.matrix(cluster_mod_exp),
                 fontsize = 8,
                 scale = "column",
                 color = colorRampPalette(c("#4d2d87","#7c55a3", "white", "#ed9921", "#895d25"))(n = 1000),
                 annotation_legend = F,
                 annotation_row = annot_name,
                 annotation_col = annot_module,
                 annotation_colors = annot_col,
                 border_color = NA)

celltype_order <- htmp$tree_row$labels[htmp$tree_row$order]
module_order <- htmp$tree_col$labels[htmp$tree_col$order]
```

# scaled module expression

```{r scaled-avg-exp-toplot}
toplot_avg_exp <- cluster_mod_exp %>% 
  rownames_to_column("sample_celltype") %>%
  gather(key = "module", value = "avg_clust_mod_exp", 2:11) %>% 
  mutate(key = paste(sample_celltype, module, sep = "/")) %>% 
  left_join(detect_rate, by = "key") %>% 
  mutate(sample_celltype = factor(sample_celltype, levels = rev(celltype_order))) %>% 
  mutate(module = factor(module, levels = rev(module_order)))

# calculate module mean expresion and sd
scale_factors <- toplot_avg_exp %>%
  group_by(module) %>% 
  reframe(sd = sd(avg_clust_mod_exp),
          mean = mean(avg_clust_mod_exp))

toplot_avg_exp_scaled <- toplot_avg_exp %>% 
  left_join(scale_factors, by = "module") %>% 
  mutate(exp_scaled = (avg_clust_mod_exp - mean)/sd) %>% 
  # cap scaled values at abs(5)
  mutate(exp_scaled_5 = case_when(
    exp_scaled > 5 ~ 5,
    exp_scaled < -5 ~ -5,
    TRUE ~ exp_scaled
  ))
```

# dotplot

```{r, fig.height= 10, fig.width=6}

palette <- colorRampPalette(c("#4d2d87","#7c55a3", "white", "#ed9921", "#895d25"))(n = 1000)
palette <- palette[floor(500+min(toplot_avg_exp_scaled$exp_scaled_5)*100):1000]


dotplot <- ggplot(toplot_avg_exp_scaled, 
       aes(x = module, 
           y = sample_celltype,
           size = detection_rate,
           color = exp_scaled_5)) +
  geom_point() +
  scale_size(range = c(0.5,3.5)) +
  scale_color_gradientn(colors = palette) +
  theme_cowplot() +
  ggtitle("scaled avg_module_expression (values capped at abs(5)") +
  theme(axis.text.x = element_text(angle = 90, vjust = 0.5, hjust = 1),
        axis.text = element_text(size = 7))

dotplot
```

```{r export}
module_order <- c("lightgreen","darkred","turquoise","brown","cyan",
                  "purple","midnightblue","lightcyan","blue","salmon",
                  "lightyellow","pink","royalblue","darkgreen","tan",
                  "yellow","grey60","greenyellow","green","red",
                  "magenta","black")

pdf("~/spinal_cord_paper/figures/Fig_1_devel_module_elbowplots.pdf", width = 8, height = 6)
grid.arrange(grobs = elbowplots[module_order], ncol = 6, nrow = 4)
dev.off()

pdf("~/spinal_cord_paper/figures/Fig_1_devel_module_fitted_elbowplots.pdf", width = 12, height = 4)
grid.arrange(grobs = fit_elbow[modules_to_plot], ncol = 5, nrow = 2)
grid.arrange(grobs = elbowplots[rev(module_order)], ncol = 11, nrow = 2)
dev.off()



pdf("~/spinal_cord_paper/figures/Fig_1_TF_devel_module_correlation.pdf", width = 8, height = 6)
ps_cor_boxplot
sm_cor_boxplot
dev.off()

pdf("~/spinal_cord_paper/figures/Fig_1_devel_module_v_clusters_dotplot.pdf", width = 5.5, height = 10)
htmp
dotplot
dev.off()

write.csv(ps_cor_long, "~/spinal_cord_paper/output/module_devel_TF_corr_pearson.csv")
write.csv(sm_cor_long, "~/spinal_cord_paper/output/module_devel_TF_corr_spearman.csv")
```


```{r}
# Date and time of Rendering
Sys.time()

sessionInfo()
```
