---
title: "Motor neuron counting"
author: "Fabio Sacher"
date: "17.06.2024"
data:
output:
  html_document:
    df_print: paged
    toc: TRUE
    toc_float: TRUE
  html_notebook:
    fig_height: 7
    fig_width: 8
editor_options:
  chunk_output_type: inline
---


```{r libraries}
library(dplyr)
library(tibble)
library(stringr)
library(magrittr)
library(tidyr)
library(ggplot2)
library(ggpubr)
library(patchwork)
```

# counting data

Load the motor neuron counting data.

```{r load-files}
files <- list.files("~/spinal_cord_paper/data/MN_counting/", pattern = ".txt")

tables <- list()

for (i in seq(files)) {
  tables[[i]] <- read.delim(paste0("~/spinal_cord_paper/data/MN_counting/", files[i]), sep = ",")
}

names(tables) <- files
```

# data wrangling

Combine the tables into a single data frame and bring it into long format. Also define a split to separate sections into rostral and caudal. 
The value used is the median of the index range by day.

```{r data-wrangling}
df <- do.call(rbind,tables) %>% 
  rownames_to_column("tmp") %>% 
  mutate(index = str_extract(tmp, "\\d{1,2}$")) %>% 
  mutate(day = str_extract(tmp, "^Day\\d{1,2}")) %>% 
  mutate(day = factor(day, levels = c("Day6", "Day8", "Day10"))) %>% 
  mutate(embryo = str_extract(tmp, "emb_[abcdef]")) %>% 
  mutate(index = as.integer(index)) %>% 
  select(ctrl, poly, index, day, embryo) %>% 
  mutate(embryo_id = paste(day, embryo, sep = "_"))

# factor levels to order the embryo ids
id_levels <- c("Day6_emb_a","Day6_emb_b","Day6_emb_c","Day6_emb_d","Day6_emb_e",
               "Day8_emb_a","Day8_emb_b","Day8_emb_c","Day8_emb_d","Day8_emb_e",
               "Day10_emb_a","Day10_emb_b","Day10_emb_c","Day10_emb_d", "Day8_emb_f")

long_df <- gather(df, "condition", "count", -index, -day, -embryo, -embryo_id) %>% 
  mutate(condition = factor(condition, levels = c("ctrl", "poly"))) %>% 
  mutate(embryo_id = factor(embryo_id, levels = id_levels))

# use median to split indeces into rostral and caudal
axis_split <- long_df %>% 
  group_by(day) %>% 
  summarise(split = median(range(index)))

splits <- axis_split$split
names(splits) <- axis_split$day

long_df <- long_df %>% 
  mutate(axis = case_when(
    day == names(splits)[1] & index < splits[[1]] ~ "caudal",
    day == names(splits)[2] & index < splits[[2]] ~ "caudal",
    day == names(splits)[3] & index < splits[[3]] ~ "caudal",
    TRUE ~ "rostral"
  ))  %>% 
  mutate(split = case_when(
    day == names(splits)[1] ~ splits[[1]],
    day == names(splits)[2] ~ splits[[2]],
    day == names(splits)[3] ~ splits[[3]]
  ))

```

# initial data plotting

Plot the raw data split by embryonic days.

1. total counts by embryo

2. counts split by condition by embryo

3. counts split by condition

```{r initial-plotting, fig.height=7, fig.width=10}
ggplot(data = long_df, aes(x = index, y = count, color = embryo)) +
  geom_point() +
  geom_smooth() +
  geom_vline(aes(xintercept = split), lty = "dashed") +
  facet_wrap("day", nrow = 3)

ggplot(data = long_df, aes(x = index, y = count, color = condition)) +
  geom_point() +
  geom_smooth() +
  geom_vline(aes(xintercept = split), lty = "dashed") +
  facet_wrap("embryo_id", nrow = 3)

ggplot(data = long_df, aes(x = index, y = count, color = condition)) +
  geom_point() +
  geom_smooth() +
  geom_vline(aes(xintercept = split), lty = "dashed")+
  facet_wrap("day", nrow = 3)

```

# data cleaning

There is an outlier in Day10, Emb_d, slide 54 with 550 nuclei. Most likely an
added 0 (typo).
Since it is almost 7X the 
mean of values from slides 49 to 59 of all embryos and conditions (79.29487) 
it is removed.

Also embryo f from day 8 is removed since it was only counted partially.

```{r cleaning}
mean(long_df$count[long_df$day == "Day10" &
                     long_df$index %in% c(49:59)],
     na.rm = TRUE)
# set the outlier to NA
long_df$count[long_df$count == 550] <- NA
# remove Day 8 embryo f
long_df <- long_df[!(long_df$embryo_id == "Day8_emb_f"),]
```

# plotting cleaned data

```{r plot-cleaned-data}
ggplot(data = long_df, aes(x = index, y = count, color = embryo)) +
  geom_point() +
  geom_smooth() +
  geom_vline(aes(xintercept = split), lty = "dashed")+
  facet_wrap("day", nrow = 1)

sup <- ggplot(data = long_df, aes(x = index, y = count, color = condition)) +
  geom_point(size = 1, alpha = 0.5) +
  geom_smooth(linewidth = 0.5) +
  geom_vline(aes(xintercept = split), lty = "dashed")+
  scale_x_reverse() +
  scale_color_manual(values = c("black", "darkgoldenrod3")) +
  facet_wrap("embryo_id", nrow = 3)

main <- ggplot(data = long_df, aes(x = index, y = count, color = condition, shape = condition)) +
  geom_point(size = 1, alpha = 0.2) +
  geom_vline(aes(xintercept = split), lty = "dashed")+
  scale_shape_manual(values = c(1, 4)) +
  scale_color_manual(values = c("black", "darkgoldenrod3")) +
  scale_x_reverse() +
  geom_smooth(method = "gam", formula = y ~ s(x, bs = "cs")) +
  facet_wrap("day", nrow = 3) + theme_bw()

sup

main

```

# barplots of total MN count

Next we plot barplots of total motor neuron count to see overal changes.

```{r data-prep-total-barplot}
nMN_df <- long_df %>% 
  unite(embryo_id, condition, col = "id", sep = "-") %>% 
  group_by(id) %>% 
  summarize(nMN = sum(count, na.rm = TRUE)) %>%
  separate(id, sep = "-", into = c("embryo_id", "cond")) %>%
  mutate(day = str_extract(embryo_id, "^Day\\d{1,2}")) %>% 
  mutate(grouping = paste(day, cond, sep = "_")) %>% 
  group_by(grouping) %>% 
  mutate(day = factor(day, levels = c("Day6", "Day8", "Day10")))

std.error <- function(x) sd(x)/sqrt(length(x))

se_df <- nMN_df %>% summarise(se = std.error(nMN))
avg_df <- nMN_df %>% summarise(mean = mean(nMN)) %>% 
  left_join(se_df, by = "grouping") %>% 
  separate(grouping, sep = "_", into = c("day", "cond")) %>% 
  mutate(day = factor(day, levels = c("Day6", "Day8", "Day10")))

# Define the top and bottom of the errorbars
limits <- aes(ymax = mean + se, ymin = mean - se)
dodge <- position_dodge2(width=0.9)

```

```{r plot-total-barplot, fig.width=3}
# total counts ctrl vs poly by day
bar_cond <- ggplot(data = avg_df,
       aes(x = day,
           y = mean,
           color = cond)) + 
  geom_bar(position = dodge, stat="identity", fill = NA) +
  geom_linerange(limits, position=dodge) +
  scale_color_manual(values = c("black", "darkgoldenrod3")) +
  facet_wrap("day", nrow = 3, scales = "free_x") +
  theme_bw()

bar_cond
```

## run tests between groups

```{r tests-between-groups}

compare_means(nMN ~ cond, data = nMN_df %>% ungroup() %>% filter(day == "Day6"))
compare_means(nMN ~ cond, data = nMN_df %>% ungroup() %>% filter(day == "Day8"))
compare_means(nMN ~ cond, data = nMN_df %>% ungroup() %>% filter(day == "Day10"))

compare_means(nMN ~ day,
              data = nMN_df %>%
                    ungroup() %>% 
                    filter(cond == "ctrl"))

my_comparisons <- list(c("Day6", "Day8"), c("Day8", "Day10"), c("Day6", "Day10"))
```

```{r, fig.width=5}
box_day_ctrl <- ggplot(data = nMN_df %>%
        ungroup() %>% 
        filter(cond == "ctrl"),
       aes(x = day, y = nMN, color = cond, shape = cond)) + 
  geom_boxplot() +
  scale_color_manual(values = c("black")) +
  stat_compare_means(comparisons = my_comparisons) +
  geom_point() +
  scale_shape_manual(values = c(1)) +
  ylim(4750,9050) +
  theme_bw()

box_day_poly <- ggplot(data = nMN_df %>%
        ungroup() %>% 
        filter(cond == "poly"),
       aes(x = day, y = nMN, color = cond, shape = cond)) + 
  geom_boxplot() +
  scale_color_manual(values = c("darkgoldenrod3")) +
  stat_compare_means(comparisons = my_comparisons) +
  geom_point() +
  scale_shape_manual(values = c(4)) +
  ylim(4750,9050) +
  theme_bw()

box_day <- box_day_ctrl + 
  box_day_poly + 
  plot_layout(guides = "collect") + 
  plot_annotation( title = "MN counts across devel")

box_day
```

# barplots rostral vs caudal

```{r rost-vs-caud-barplot, fig.width=5}
axis_nMN <- long_df %>% 
  mutate(axis = case_when(
    day == names(splits)[1] & index < splits[[1]] ~ "caudal",
    day == names(splits)[2] & index < splits[[2]] ~ "caudal",
    day == names(splits)[3] & index < splits[[3]] ~ "caudal",
    TRUE ~ "rostral"
  )) %>% 
  unite(embryo_id, condition, axis, col = "id", sep = "-") %>% 
  group_by(id) %>% 
  summarize(nMN = sum(count, na.rm = TRUE)) %>%
  separate(id, sep = "-", into = c("embryo_id", "cond", "axis")) %>%
  mutate(day = str_extract(embryo_id, "^Day\\d{1,2}")) %>% 
  mutate(grouping = paste(day, cond, axis, sep = "_")) %>% 
  group_by(grouping) %>% 
  mutate(day = factor(day, levels = c("Day6", "Day8", "Day10")))

# grouping for the barplot
axis_levels <- c("Day6_ctrl_rostral",
                 "Day6_poly_rostral",
                 "Day6_ctrl_caudal",
                 "Day6_poly_caudal",
                 "Day8_ctrl_rostral",
                 "Day8_poly_rostral",
                 "Day8_ctrl_caudal",
                 "Day8_poly_caudal",
                 "Day10_ctrl_rostral",
                 "Day10_poly_rostral",
                 "Day10_ctrl_caudal",
                 "Day10_poly_caudal")

se_axis_df <- axis_nMN %>% summarise(se = std.error(nMN))
avg_axis_df <- axis_nMN %>% summarise(mean = mean(nMN)) %>% 
  left_join(se_axis_df, by = "grouping") %>% 
  separate(grouping, sep = "_", into = c("day", "cond", "axis"), remove = FALSE) %>% 
  mutate(day = factor(day, levels = c("Day6", "Day8", "Day10"))) %>% 
  mutate(grouping = factor(grouping, levels = axis_levels))

# Define the top and bottom of the errorbars
limits <- aes(ymax = mean + se, ymin = mean - se)

bar_axis <- ggplot(data = avg_axis_df,
       aes(x = grouping,
           y = mean,
           color = cond)) + 
  geom_bar(position = dodge, stat="identity", fill = NA) +
  geom_point(data = axis_nMN,
             aes(x = grouping,
                 y = nMN,
                 shape = cond)) +
  geom_linerange(limits, position=dodge) +
  scale_color_manual(values = c("black", "darkgoldenrod3")) +
  scale_shape_manual(values = c(1, 4)) +
  facet_wrap("day", nrow = 3, scales = "free_x") +
  theme_bw()

bar_axis

my_comparisons <- list(c("Day6_ctrl_rostral", "Day6_poly_rostral"),
                       c("Day6_ctrl_caudal",  "Day6_poly_caudal"),
                       c("Day8_ctrl_rostral", "Day8_poly_rostral"),
                       c("Day8_ctrl_caudal",  "Day8_poly_caudal"),
                       c("Day10_ctrl_rostral","Day10_poly_rostral"),
                       c("Day10_ctrl_caudal", "Day10_poly_caudal")
                       )

# total counts per day by condition
box_axis <- ggplot(data = axis_nMN %>%
         ungroup() %>% 
         mutate(grouping = factor(grouping, levels = axis_levels)),
       aes(x = grouping, y = nMN, color = cond)) + 
  geom_boxplot() +
  scale_color_manual(values = c("black", "darkgoldenrod3")) +
  stat_compare_means(comparisons = my_comparisons) +
  ggtitle("rostral vs caudal MN counts across devel") +
  facet_wrap("day", nrow = 3) +
  theme_bw() +
  theme(axis.text.x = element_text(angle = 45, hjust = 1))

box_axis
```

# figure export

```{r}

ggsave(
  filename = "~/spinal_cord_paper/figures/line_plot_MN_count_individual.pdf",
  width = 10, height = 6,
  plot = sup
)

ggsave(
  filename = "~/spinal_cord_paper/figures/Fig_5_line_plot_MN_count.pdf",
  width = 4, height = 6,
  plot = main
)

ggsave(
  filename = "~/spinal_cord_paper/figures/Fig_5_ctrl_vs_poly_barplot_MN_count.pdf",
  width = 3, height = 10, 
  plot = bar_cond
)

ggsave(
  filename = "~/spinal_cord_paper/figures/Supp_Fig_5_ctrl_vs_poly_boxplot_MN_count.pdf",
  width = 6, height = 5,
  plot = box_day
)

ggsave(
  filename = "~/spinal_cord_paper/figures/Fig_5_barplot_split_MN_count.pdf",
  width = 3, height = 10,
  plot = bar_axis
)

ggsave(
  filename = "~/spinal_cord_paper/figures/Fig_5_boxplot_split_MN_count.pdf",
  width = 7, height = 10,
  plot = box_axis
)

```



# Session Info

```{r Session-info}
# Date and time of Rendering
Sys.time()

sessionInfo()
```